---
title: "Final"
author: "Monique Hernandez"
date: "June 11, 2025"
format:
  html:
    toc: true
execute:
  message: false
  warning: false
---

[Github link](https://github.com/moniqueallyson/ENVS-193DS_spring-2025_final)

# Set up

```{r, message=FALSE}


library(tidyverse) # general use
library(janitor) # cleaning data frames
library(here) # file organization
library(MuMIn) # model selection
library(DHARMa) # model diagnostics
library(ggeffects) # getting model predictions
library(scales) # modifying axis labels


# read in data
sst <- read.csv(here("data", "SST_update2023.csv"))

# read in data
nest_boxes <-  read.csv(
  # file path
  here("data", "occdist.csv"))


```

# Problem 1. Research writing

## a. Transparent statistical methods

In part 1, they used a correlation test, specifically Pearson's r, since they measured the strength and direction of two variables which in this case are two continuous variables: distance from headwater(km) and annual total nitrogen load (kg year^-1^).

In part 2, thy used a one-way ANOVA test because the report compares the average nitrogen load across more than two groups: urban land, atmospheric deposition, fertilizer, wastewater treatment, and grasslands.

## b. More information needed

From Part 2:

1.  It would be helpful to do a post-hoc test like Tukey's HSD as it can determine which groups are different from each other while the ANOVA test signified that there is a difference in the average nitrogen load (kg year⁻¹) between sources (p = 0.02). The Tukey's HSD will be able to compare the groups to each other (fertilizer vs. grasslands, urban land vs. atmospheric deposition), which will determine which pairs of groups have significantly different mean nitrogen loads.

2.  It could also be helpful to use effect size, such as eta-squared (η²), which measures the proportion of variation explained in nitrogen load (kg year⁻¹) from the five different sources. This means that it quantifies the portion of variance explained by sources. The interpretation for the results can range from negligible (η² \< .01), small (.01 \< η² \< .06), medium (.06 \< η² \< .14), and large (η² \> .14).

## c. Suggestions for rewriting

Part 1: There is a relationship between distance from headwater (km) and annual total nitrogen load (kg year⁻¹), suggesting that nitrogen load varies with distance. We rejected the null hypothesis that there is no correlation between these variables (Pearson’s r = \[correlation value\], p = 0.03, α = 0.05).

Part 2: A difference in the average nitrogen load (kg year⁻¹) in the San Joaquin River Delta was found between the five sources (urban land, atmospheric deposition, fertilizer, wastewater treatment, and grasslands), with an effect size (η² = effect size value) indicating that the proportion of variance explained by source type. This suggests that these sources contribute to variation in nitrogen load in the river delta. (One-way ANOVA; F(4, df₂) = F-value, p = 0.02, α = 0.05)

# Problem 2. Data visualization

## a. Cleaning and summarizing

```{r}

sst_clean <- sst |>
  mutate(date = ymd(date), # convert date column to Date format
    year = year(date),       # Extract year from date
    month = month(date, label = TRUE)  # Extract month as a labeled factor
  ) |>
  filter(year == c("2018", "2019", "2020", "2021", "2022", "2023"))|>  # keep years from 2018 to 2023
  select(year, month, temp) |> #select columns year, month, and temp
  group_by(year, month) |>               # Group by year and month
  summarize(
    mean_monthly_sst = mean(temp, na.rm = TRUE)  # Calculate mean temp
  ) |>
  ungroup()  # remove grouping

# display 5 random rows
sst_clean |> slice_sample(n = 5)

# show the structure of the cleaned data
str(sst_clean)

```

## b. Visualize the data

```{r}

sst_visual <- ggplot(data = sst_clean,  # new data called sst_visual
       aes(x = month, # month on x-axis
           y = mean_monthly_sst, # year on y-axis
           group = factor(year), # group by year
           color = factor(year))) + # color by year
  geom_line() +                          
  geom_point() +                         
  scale_color_manual(name = "Year", # manually assign colors by group
    values = c("2018" = "#E6E6FA",
       "2019" = "#D8BFD8",
       "2020" = "#DDA0DD",
       "2021" = "#DA70D6",
       "2022" = "#BA55D3",
       "2023" = "#8A2BE2")) +
  labs(x = "Month", # label the title on x-axis
    y = "Mean Monthly Sea Surface Temperature (°C)") + # label y-axis title
  theme_minimal() +                             
  theme(legend.position = c(0.15, 0.65), # position legend lower left
        plot.background = element_rect(fill = "white"), # background white
        panel.grid.major = element_blank(),   # remove major grid lines
        panel.grid.minor = element_blank()) #remove minor grid lines

sst_visual

```

# Problem 3. Data analysis

```{r}
#| include: false
#| label: Cleaning next_boxes data

nest_clean <- nest_boxes |> # clean data
  clean_names() |> # clean column names
  mutate(season = as.factor(season)) # make season into a factor

nest_explore <- nest_boxes |>
  clean_names() |> # clean columns and replace with _ 
  mutate(bin = cut(edge_distance, breaks = 10)) |>
  group_by(bin) |>
  mutate(prop_occupied = mean(sp),
            mid = mean(edge_distance)) |>
  ungroup() |> 
  mutate(season = factor(season))


# plot occupancy proportion vs edge distance bins
ggplot(data = nest_explore,
       aes(x = mid,
           y = prop_occupied)) +
  geom_point() +
  geom_smooth(method = "lm", se = FALSE) 


```

## a. Response variable

The columns cs, sp, and tm, abbreviated for Common Starling, Swift Parrot, and Tree Martin, contain 1's and 0's that indicate whether the nest box was occupied by the respective species (1 = occupied by that species, 0 = not). The column e, abbreviated for “empty,” uses 1 to indicate the box was empty and 0 to indicate it was occupied by any species.

## b. Purpose of study

Swift Parrots are a critically endangered species and the target for habitat restoration efforts, while Common Starlings and Tree Martins are competitor species that exploit these nest boxes. Unlike the competitors, which benefited more from established boxes, Swift Parrots use cavities only during specific breeding periods and do not return to the same nesting sites from previous years.

## c. Difference in “seasons”

Due to a mass flowering event, the summers of 2016 and 2019 were the focus for studying Swift Parrot breeding habits and nest box usage. The boxes were first deployed in 2016, and by 2019, competing species, such as Tree Martin and Common Starlings, had started using them more often during their own nesting periods, while Swift Parrots were only present during the flowering years and absent in-between.

## d. Table of models

| Model number | Season | Distance to forest edge | Predictor list           |
|:------------:|:------:|:-----------------------:|--------------------------|
|      1       |        |                         | No predictors (null)     |
|      2       |   X    |            X            | Season + Distance (full) |
|      3       |   X    |                         | Season only              |
|      4       |        |            X            | Distance only            |

## e. Run the models

```{r}


model1 <-  glm(sp ~ 1, # null model
                   data = nest_clean,
                   family = "binomial") 

model2 <-  glm(sp ~ season + edge_distance, # season + distance
             data = nest_clean,
             family = "binomial")

model3 <- glm(sp ~ season, # season
            data = nest_clean,
            family = "binomial")

model4 <-  glm(sp ~ edge_distance, # distance
             data = nest_clean,
             family = "binomial")


```

## f. Check the diagnostics

```{r}

# simulate residuals for each model
residuals1 <- simulateResiduals(model1)
residuals2 <- simulateResiduals(model2)
residuals3 <- simulateResiduals(model3)
residuals4 <- simulateResiduals(model4)

# plotting diagnostics for each model
plot(residuals1, main = "Model1: Null Model")
plot(residuals2, main = "Model 2: Saturated Model")
plot(residuals3, main = "Model 3: Season Only Model")
plot(residuals4, main = "Model 4: Distance Only Model")



```

## g. Select the best model

```{r}

AICc(model1, # Akaike’s Information Criterion choosing best model
     model2,
     model3,
     model4) |>
 
  # arranging output in descending order of AIC
  arrange(AICc)

```

The best model as determined by Akaike’s Information Criterion (AIC) is the saturated model on season (year) and edge distance to the forest for Swift Parrot's occupancy. This model suggests that the year the nesting occurred as well as the distance were part of the influence whether the Swift Parrots used the nest boxes.

## h. Visualize the model predictions

```{r}

# make model prediction as new object
model_prediction <- ggpredict(model2, 
                              terms = c("edge_distance [all]", "season")) # breeding season from 2016 and 2019

# convert predictions to a data frame
model_df <- as.data.frame(model_prediction) |>
  rename(Season = group)  # rename group column 

# base layer: gg plot
ggplot(model_df, 
       aes(x = x,  
           y = predicted, 
           color = Season,
           fill = Season)) +
  geom_ribbon(aes # add geom ribbon for confidence interval @ 95%
              (ymin = conf.low,
                ymax = conf.high),
              alpha = 0.25, color = NA) +
  geom_line() +
  geom_point(size = 1.2) +
  geom_jitter(data = nest_explore, 
              aes(x = edge_distance, y = sp, color = season),
              inherit.aes = FALSE,
              height = 0.02, 
              alpha = 0.35,
              size = 1,
              shape = 2) +
  scale_color_manual(values = c("2016" = "violet", "2019" = "purple")) +
  scale_fill_manual(values = c("2016" = "violet", "2019" = "purple")) +
  scale_y_continuous(labels = percent_format(accuracy = 1)) +
  labs(
    x = "Distance from Forest Edge (m)",
    y = "Likelihood of Swift Parrot Occupancy",
    title = "Predicting Swift Parrot Nest Box Use by Season"
  ) +

  theme_bw() +
  theme(
    panel.grid.major = element_blank(), # remove grind lines
    panel.grid.minor = element_blank() # remove minor grind lines
  )

```

## i. Write a caption for your figure.

**Figure 1. Swift Parrots have higher likelihood for box occupancy when close to forest edge:** Extracted data from the DRYAD (Saunders DL, Mawson PR, Dawson R,2020 DOI: 10.1111/rec.13319). Lines show predictions from a saturated generalized linear model with shaded ribbons indicating 95% confidence intervals, differentiated by season through the use of violet for 2016 and purple for 2019. The circular points represent the likelihood of a Swift Parrot occupancy in a nest box according to the distance from the forest edge, while triangular points represent the binary data representing occupancy.

## j. Calculate model predictions

```{r}
#| echo: true

# predict for 0 m from forest edge with 95% CI
ggpredict(model2, terms = c("edge_distance [0]", "season"))


# predict for 900 m from forest edge with 95% CI
ggpredict(model2, terms = c("edge_distance [900]", "season"))


```

## k. Interpret your results

Utilizing the figure and prediction results, the Swift Parrot occupancy is highest near the forest edge (0 m) with a 48% occupancy predicted in 2016 and 30% in 2019. Comparing this to 900 m from the forest edge, the predicted occupancy lowers to 12% in 2016 and 6% in 2019 which solidify the inverse relationship between Swift Parrot occupancy and forest distance. As stated in the article's results. from the twenty boxes used in 2016 only five were re-used in 2019. This may be due to the fact that Swift Parrots rarely as stated breed in the same location as the previous year.

# Problem 4. Affective and exploratory visualizations

## a. Comparing visualizations

It's different because the exploratory visualization shows the jitter plot between two groups and the difference between them while affective shows a plain picture. The latter was not representing the data in a clear way through an artistic expression as it was created initially to do so.

The similarities I see between the two is the up and down motion from the affective visualization referencing the range of chronic pain into a wave or rhythmic motion just as the exploratory has two values being high and low values in th jitter plot.

The relationships are different as the affective doesn't allow for the information to come across neatly like it does for the jitter plot. This could have helped provide more insight by depicting the variables as in when a yoga session was done or not and how it related to the chronic pain felt through that day,

The feedback received through written note was what I mentioned previously how there was not a clear data being represented through the affective visualization as it was done through the exploratory option. At the moment, I would try to implement changes by having perhaps a mountain as then the ocean, despite its turbulent waves, could represent the low pain days when yoga was practiced (a journey itself) versus the heights of the mountains depicting days where I did not practice yoga and experienced high pain, This new ideal attempt reminds me of the famous Japanese painting titled "The Wave".

## b. Sharing your affective visualization

Workshop 10, I was not present to share my affective visualization though after seeing other classmates I see how I would have changed mine to be more technical rather than artistic. I took the artist approach to visualization too literally.
